# Data Manipulation and Aggregation {#sec-manip}

```{mermaid}
flowchart LR
dt[data.table]
sub[Subsetting Observations<br>and Variables]
am[Adding/Modifying Variables]
rec[Recode] --> hrec[R Expressions<br>factor<br>fcase<br>score.binary<br>table look-up]
res[Reshape] --> lw[Long to Wide<br>dcast] & wl[Wide to Long<br>melt] 
```

The [`data.table`](https://hbiostat.org/R/data.table) package provides a concise, consistent syntax for managing simple and complex data manipulation tasks, and it is extremely efficient for large datasets.  One of the best organized tutorials is [this](https://riptutorial.com/data-table), and a cheatsheet for data transformation is [here](https://raw.githubusercontent.com/rstudio/cheatsheets/master/datatable.pdf).  Userful `data.table` vignettes are [here](https://cran.r-project.org/web/packages/data.table/vignettes). A master cheatsheet for `data.table` is [here](https://rdrr.io/cran/data.table/man/data.table.html) from which the general syntax below is taken, where `DT` represents a data table.

```
   DT[ i,  j,  by ] # + extra arguments
        |   |   |
        |   |    -------> grouped by what?
        |    -------> what to do?
         ---> on which rows?
```

The following chart may help one to conceptualize the `data.table` framework.  To enter into the environment of a data table means to be able to address its columns directly without having, for example, the `DT` prefix.

```{mermaid}
flowchart TD
DTO["DT ["]
E["Enter Environment<br>of DT<br>DT[rows,columns,by/on]"]
DTO --> E
R[Rows To Fetch<br>Or Change]
C["Columns to Fetch<br><hr>Or Create/Redefine<br>with :="]
OB["by (grouping)<br><hr>on (matching)"]
E --> R & C & OB
R & C & OB --> DTC["]"] --> L[Leave<br>Environment]
L --> N["No :=<br><br>Print/Plot Output<br>or Assign Result To<br>New Data Table"]
L --> IP[":=<br><br>Change DT in Place"]
```

Data tables are also data frames so work on any method handling data frames.  But data tables do not contain `rownames`.

Several `data.table` examples follow.  I like to hold the current dataset in `d` to save typing.  Some basic operations on data tables are:

```{r eval=FALSE}
d[2]                  # print 2nd row
d[2:4]                # print rows 2,3,4
d[month == 'November']# fetch rows for November
d[month %like%  'mb'] # fetch rows with 'mb' in month
d[month %ilike% 'mb'] # fetch rows with 'mb' in month ignoring case
d[y > 2 & z > 3]      # rows satisfying conditions
d[, age]              # retrieve one variable
d[, .(age, gender)]   # make a new table with two variables
i <- 2; d[, ..i]      # get column 2
v <- 'age'; d[, ..v]  # get variable named by contents of v
d[, (v)]              # another way; these return data tables
d[.N]                 # last row
d[, .N, keyby=age]    # number of rows for each age, sorted
d[1:10, bhr:pkhr]     # first 10 rows, variables bhr - pkhr
d[1:10, !(bhr:pkhr)]  # all but those variables
d[, 2:4]              # get columns 2-4
d[, {...}]            # run any number of lines of code in ...
                      # that reference variables inside d
```

`data.table` does many of its operations _by reference_ to avoid the overhead of having multiple copies of data tables.  This idea carries over to _apparent_ copies of data tables.  Here is an example.

```{r dfcopy}
require(Hmisc)
require(data.table)
knit_print <- knitr::normal_print # make knitr standard print output (not kable)
a <- data.table(x1=1:3, x2=letters[1:3])
a
b <- a                  # no copy; b is just a pointer to a
b[, x2 := toupper(x2)]  # changes a
a
a <- data.table(x1=1:3, x2=letters[1:3])
a2 <- copy(a)           # fresh copy with its own memory space
a2[, x2 := toupper(x2)] # doesn't change a
a
```

## Analyzing Selected Variables and Subsets
```{r dtsimple,results='asis'}
getHdata(stressEcho)
d <- stressEcho
setDT(d)
d[, html(describe(age))]
d[, html(describe(~ age + gender))]
d[gender == 'female', html(describe(age))]   # analyze age for females
html(describe(d[, .(age, gender)], 'Age and Gender Stats'))
```

```{r dtsimple2}
# Separate analysis by female, male.  Use keyby instead of by to sort the usual way.
d[, print(describe(age, descript=gender)), by=gender]

# Compute mean and median age by gender
d[, .(Mean=mean(age), Median=median(age)), by=gender]
# To create a new subset
w <- d[gender == 'female' & age < 70, ]
```

## Adding, Changing, and Removing Variables {#sec-manip-acr}

With `data.table` you can create a new data table with added variables, or you can add or redefine variables in a data table in place.  The latter has major speed and memory efficiency implications when processing massive data tables.  Here `d` refers to a different data table from the one used above.

```{r eval=FALSE}
# Rename a variable
setnames(d, c('gender', 'height'),
            c(  'sex',      'ht'))
# Easier:
setnames(d, .q(gender, height),
            .q(   sex,     ht))

# Or
ren <- .q(gender=sex, height=ht)
setnames(d, names(ren), ren)

# Or
rename <- function(x, n) setnames(x, names(n), n)
rename(d, .q(gender=sex, height=ht))

# For all variables having baseline_ at the start of their names, remove it
n <- names(d)
setnames(d, n, sub('^baseline_', '', n))   # ^ = at start; use $ for at end

# For all variables having baseline_ at the start of their names, remove it
# and add Baseline to start of variable label
n    <- names(d)
n    <- n[grepl('^baseline_', n)]
ren  <- sub('^baseline_', '', n); names(ren) <- n
# Fetch vector of labels for variables whose names start with baseline_
labs <- sapply(d, label)[n]   # label() result is "" if no label
labs <- paste('Baseline', labs)
d    <- updata(d, rename=ren, labels=labs)

# Change all names to lower case
n <- names(d)
setnames(d, n, tolower(n))

# Abbreviate names of all variables longer than 10 characters
# abbreviate() will ensure that all names are unique
n <- names(d)
setnames(d, n, abbreviate(n, minlength=10))

# For any variables having [...] or (...) in their labels, assume these
# are units of measurement and move them from the label to the units
# attribute
d <- upData(d, moveUnits=TRUE)

# Add two new variables, storing
result in a new data table
z <- d[, .(bmi=wt / ht ^ 2, bsa=0.016667 * sqrt(wt * ht))]

# Add one new variable in place
d[, bmi := wt / ht ^ 2]

# Add two new variables in place
d[, `:=`(bmi = wt / ht ^ 2, bsa=0.016667 * sqrt(wt * ht))]
d[, .q(bmi, bsa) := .(wt / ht ^ 2, 0.016667 * sqrt(wt * ht))]

# Compute something requiring a different formula for different types
# of observations
d[, htAdj := ifelse(sex == 'male', ht, ht * 1.07)]  # better" use fifelse in data.table
d[, htAdj := ht * ifelse(sex == 'male', 1, 1.07)]
d[, htAdj := (sex == 'male') * ht + (sex == 'female') * ht * 1.07]
d[sex == 'male',   htAdj := ht]
d[sex == 'female', htAdj := ht * 1.07]
d[, htAdj := fcase(sex == 'male',   ht,          # fcase is in dta.table
                   sex == 'female', ht * 1.07)]
d[, htAdj := fcase(sex = 'female', ht * 1.07, default = ht)]

# Add label & optional units (better to use upData which works on data tables)
adlab <- function(x, lab, un='') {
  label(x) <- lab
  if(un != '') units(x) <- un
  x
}
d[, maxhr := adlab(maxhr, 'Maximum Heart Rate', '/m')]

# Delete a variable (or use upData)
d[, bsa := NULL]

# Delete two variables
d[, `:=`(bsa=NULL, bmi=NULL)]
d[, .q(bsa, bmi) := NULL]

# Delete a variable, but not in place (e.g, just print narrower table)
d[, ! 'bsa']
```

::: {.column-margin}
See [this](https://stackoverflow.com/questions/9202413) for a comprehensive description of ways to remove columns using `data.table`.
:::

## Recoding Variables

```{r, eval=FALSE}
# Group levels a1, a2 as a & b1,b2,b3 as b
d[, x := factor(x, .q(a1,a2,b1,b2,b3),
                   .q( a, a, b, b, b))]
# Regroup an existing factor variable
levels(d$x) <- list(a=.q(a1,a2), b=.q(b1,b2,b3))
# or
d <- upData(d, levels=list(x=list(a=.q(a1,a2), b=.q(b1,b2,b3))))
# Or manipulate character strings
d[, x := substring(x, 1, 1)]   # replace x with first character of levels
# or
levels(d$x) <- substring(levels(d$x), 1, 1)

# Recode a numeric variable with values 0, 1, 2, 3, 4 to 0, 1, 1, 1, 2
d[, x := 1 * (x %in% 1:3) + 2 * (x == 4)]
d[, x := fcase(x %in% 1:3, 1,          # fcase is in data.table
               x == 4,     2)]
d[, x := fcase(x %between% c(1,3), 1,
               x    ==       4,    2)]  # %between% is in data.table

# Recode a series of conditions to a factor variable whose value is taken
# from the last condition that is TRUE using Hmisc::score.binary
# Result is a factor variable unless you add retfactor=FALSE
d[, x := score.binary(x1 == 'symptomatic',
                      x2 %in% .q(stroke, MI),
                      death)]
# Same but code with numeric points
d[, x := score.binary(x1 == 'symptomatic',
                      x2 %in% .q(stroke, MI),
                      death,  # TRUE/FALSE or 1/0 variable
                      points=c(1,2,10))]
# Or just reverse the conditions and use fcase which stops at the first
# condition met
d[, x := fcase(death,                  'death',        # takes precedence
               x2 %in% .q(stroke, MI), 'stroke/MI',    # takes next precedence
               x1 == 'symptomatic',    'symptomatic',
               default =               'none')]

# Recode from one set of character strings to another using named vector
# A named vector can be subscripted with character strings as well as integers
states <- c(AL='Alabama', AK='Alaska', AZ='Arizona', ...)
# Could also do:
#  states <- .q(AL=Alabama, AK=Alaska, AZ=Arizona, ..., NM='New Mexico', ...) 
# or do a merge for table lookup (see later)
d[, State := states[state]]
# states are unique, state can have duplicates and all are recoded
d[, State := fcase(state == 'AL', 'Alabama',  state='AK', 'Alaska', ...)]

# Recode from integers 1, 2, ..., to character strings
labs <- .q(elephant, giraffe, dog, cat)
d[, x := labs[x]]

# Recode from character strings to integers
d[, x := match(x, labs)]
d[, x := fcase(x == 'elephant', 1, 
               x == 'giraffe',  2,
               x == 'dog',      3,
               x == 'cat',      4)]
```

As an example of more complex hierarchical recoding let's define codes in a nested `list`.

```{r hrecode}
a <- list(plant =
            list(vegetable = .q(spinach, lettuce, potato),
                 fruit     = .q(apple, orange, banana, pear)), 
          animal =
            list(domestic  = .q(dog, cat, horse),
                 wild      = .q(giraffe, elephant, lion, tiger)) )
a
a <- unlist(a)
a
# Pick names of unlist'ed elements apart to define kingdom and type
n <- sub('[0-9]*$', '', names(a))  # remove sequence numbers from ends of names
# Names are of the form kingdom.type; split at .
s       <- strsplit(n, '.', fixed=TRUE) 
kingdom <- sapply(s, function(x) x[1])
type    <- sapply(s, function(x) x[2])
# or:   (note \\. is escaped . meaning not to use as wild card)
#        .* = wild card: any number of characters
# kingdom <- sub('\\..*', '', n)  # in xxx.yyy remove .yyy
# type    <- sub('.*\\.', '', n)  # in xxx.yyy remove xxx.
names(kingdom) <- names(type) <- a
w <- data.table(kingdom, type, item=a, key=c('kingdom', 'item'))
w
# Example table lookups
cat(kingdom['dog'], ':', type['dog'], '\n')
kingdom[.q(dog, cat, spinach)]
type   [.q(dog, cat, giraffe, spinach)]
# But what if there is a plant named the same as an animal?
# Then look up on two keys
w[.('animal', 'lion'), type]
```

## Operations on Multiple Data Tables {#sec-manip-mult}

Consider an [example](https://stackoverflow.com/questions/72854617) where two data tables have almost the same columns, and rows are identified uniquely using the `id` variable.  We wish to use non-`NA` values in the second data table to fill in `NA` values in the corresponding positions in the first data table.  The solution is simple, using the `data.table` `rbind` function to combine the rows, then for every variable in the combined data table taking the first non-`NA` value for a given `id`.

```{r}
a <- data.table(id = c(1, 2, 3), v1 = c("w", "x", NA), v2 = c("a", NA, "c"),
                  v3 = 7:9)
b <- data.table(id = c(2, 3, 4), v1 = c(NA, "y", "z"), v2 = c("b", "c", NA))
w <- rbind(a, b, fill=TRUE)  # fill creates NAs for b's v3
a
b
w
g <- function(x) x[! is.na(x)][1]  # first non-NA value in vector x
u <- w[,             lapply(.SD, g), by = id]
#                    |      |    |        |
#   for every variable      |    |        |
#  in current id's data table    |        |
#               run the g function        |
#               id variable defining groups
u
```

## Reshaping Data {#sec-manip-reshape}

To reshape data from long to wide format, take the `L` data table above as an example.

```{r l2w}
L <- data.table(id  = c(2, 2, 2, 3, 3, 3, 3, 4, 4, 5, 5, 5),
                day = c(1, 2, 3, 1, 2, 3, 4, 1, 2, 1, 2, 3),
                y    =  1 : 12, key='id')
L
w <- dcast(L, id ~ day, value.var='y')
w
# Now reverse the procedure  (see also meltData in reptools)
m <- melt(w, id.vars='id', variable.name='day', value.name='y')
m
setkey(m, id, day)   # sorts
m[! is.na(y)]
```