# Merging Data

```{mermaid}
flowchart LR
SM[Merge One Record/Subject Data Table With Longitudinal Data]
HM[Hmisc::Merge<br>Monitors Merging Process]
tlu[Merging for Table Look-up]
```

Consider a baseline dataset `b` and a longitudinal dataset `L`, with subject ID of `id`. [For more information see [this](https://hbiostat.org/R/data.table), [this](https://rdrr.io/cran/data.table/man/data.table.html), [this](https://medium.com/analytics-vidhya/r-data-table-joins-48f00b46ce29), [this](https://stackoverflow.com/questions/15170741) and [this](https://stackoverflow.com/questions/13493124).  To merge any number of datasets at once and obtain a printed report of how the merge went, use the `Hmisc` `Merge` function.]{.aside}

```{r two}
require(Hmisc)
require(data.table)
b <- data.table(id=1:4, age=c(21, 28, 32, 23), key='id')
L <- data.table(id  = c(2, 2, 2, 3, 3, 3, 3, 4, 4, 5, 5, 5),
                day = c(1, 2, 3, 1, 2, 3, 4, 1, 2, 1, 2, 3),
                y    =  1 : 12, key='id')
b
L
# Merge b and L to look up baseline age and associate it with all follow-ups
b[L, on=.(id)]   # Keeps all ids in L (left inner join)
L[b, on=.(id)]   # Keeps all ids in b (right inner join)
L[b, on=.(id), nomatch=NULL]  # Keeps only ids in both b and L (right outer join)
uid <- unique(c(b[, id], L[, id]))
L[b[.(uid), on=.(id)]]         # Keeps ids in either b or c (full outer join)
merge(b, L, by='id', all=TRUE) # also full outer join; calls merge.data.table
# Use the Hmisc Merge function which reports on how the merge went
# Merge operates on both data.tables and data.frames and can handle
# any number of them.  In details printed by Merge below, #1
# refers to the first data table (b).
Merge(b, L, id= ~ id)
```

For very large data tables, giving the data tables _keys_ will speed execution, e.g.:

```
setkey(d, id)
setkey(d, state, city)
```

Join/merge can be used for data lookups:

```{r lookup}
s <- data.table(st=.q(AL, AK, AZ, CA, OK), y=5:1)
stateAbbrevs <- data.table(state=state.abb, State=state.name)
s[stateAbbrevs, , on=.(st=state), nomatch=NULL]
```