# Computing Summary Statistics {#sec-sstats}

```{mermaid}
flowchart LR
oav[Statistical Summaries of All Variables]
osv[Summaries of Variable Subsets]
ofun[Summaries Using Functions<br>Returning Multi-dimensional Results]
ms["Marginal Summaries<br>(e.g. Subtotals)"]
```

Many applications can use the automatically created `data.table` object `.SD` which stands for the data table for the current group being processed.  If `.SDcols` were not specified, all variables would be attempted to be analyzed.  Specify a vector of variable names as `.SDcols` to restrict the analysis.  If there were no `by` variable(s), `.SD` stands for the whole data table.

```{r ss}
require(Hmisc)
require(data.table)
getHdata(stressEcho)
d <- stressEcho
setDT(d)
# Compute the number of distinct values for all variables
nd <- function(x) length(unique(x))
d[, sapply(.SD, nd)]
# Same but only for variables whose names contain hx and either D or M
d[, sapply(.SD, nd), .SDcols=patterns('hx', 'D|M')]

# Compute means on all numeric variables
mn <- function(x) mean(x, na.rm=TRUE)
d[, lapply(.SD, mn), .SDcols=is.numeric]
# Compute means on all numeric non-binary variables
nnb <- function(x) is.numeric(x) && length(unique(x)) > 2
d[, lapply(.SD, mn), .SDcols=nnb]

# Print frequency tables of all categorical variables with > 2 levels
cmult <- function(x) ! is.numeric(x) && length(unique(x)) > 2
tab <- function(x) {
  z <- table(x, useNA='ifany')
  paste(paste0(names(z), ': ', z), collapse=', ')
}
d[, lapply(.SD, tab), .SDcols=cmult]
```

Tabulate all variables having between 3 and 10 distinct values and create a side effect when `data.table` is running that makes the summarization function `tab` store all values and frequencies in a growing list `Z` so that `kable` can render a `markdown` table after we pad columns to the maximum length of all columns (maximum number of distinct values). 

```{r diffreq}
# Using <<- makes data.table have a side effect of augmenting Z and
# Align in the global environment
tab <- function(x) {
  z <- table(x, useNA='ifany')
  i <- length(Z)
  Z[[i+1]] <<- names(z)
  Z[[i+2]] <<- as.vector(z)
  Align <<- c(Align, if(is.numeric(x)) 'r' else 'l', 'r')
  length(z)
}
discr <- function(x) { i <- length(unique(x)); i > 2 & i < 11 }
#                                       or i %between% c(2,11)
Z    <- list(); Align <- character(0)
w    <- d[, lapply(.SD, tab), .SDcols=discr]
maxl <- max(w)
# Pad shorter vectors with blanks
Z <- lapply(Z, function(x) c(x, rep('', maxl - length(x))))
Z <- do.call('cbind', Z)  # combine all into columns of a matrix
colnames(Z) <- rep(names(w), each=2)
colnames(Z)[seq(2, ncol(Z), by=2)] <- 'Freq'
knitr::kable(Z, align=Align)
```

A better approach is to let the `kables` function put together a series of separate `markdown` tables of different sizes.  By using the "updating `Z` in the global environment" side effect we are able to let `data.table` output any type of objects of non-conformable dimensions over variables (such as frequency tabulations).

```{r tabalt}
tab <- function(x) {
  z <- table(x, useNA='ifany')
  i <- length(Z)
  w <- matrix(cbind(names(z), z), ncol=2,
              dimnames=list(NULL, c(vnames[i+1], 'Freq')))
  Z[[i+1]] <<- knitr::kable(w, align=c(if(is.numeric(x)) 'r' else 'l', 'r'))
  length(z)
}
discr <- function(x) { i <- length(unique(x)); i > 2 & i < 11 }
Z      <- list()
vnames <- names(d[, .SD, .SDcols=discr])
w      <- d[, lapply(.SD, tab), .SDcols=discr]
knitr::kables(Z)
```

Use a similar side-effect approach to get separate `html` `describe` output by `gender`.

```{r bydesc,results='asis'}
g <- function(x, by) {
  Z[[length(Z) + 1]] <<- describe(x, descript=paste('age for', by))
  by
}
Z <- list()
by <- d[, g(age, gender), by=gender]
# Make Z look like describe() output for multiple variables
class(Z) <- 'describe'
attr(Z, 'dimensions') <- c(nrow(d), nrow(by))
attr(Z, 'descript') <- 'Age by Gender'
html(Z)
```

```{r morestats}
# Compute a 1-valued statistic on multiple variables, by cross-classification
# of two variables.  Do this on a subset.  .SDcols=a:b uses variables in order
# Use keyby instead of by to order output the usual way
d[age < 70, lapply(.SD, mean), keyby=.(gender, newMI), .SDcols=pkhr:dp]

# Compute multiple statistics on one variable
# Note: .N is a special variable: count of obs for current group
d[, .(Max=max(bhr), Min=min(bhr), Mean=mean(bhr), N=.N), by=.(gender, newMI)]
# Same result another way
g <- function(x) list(Max=max(x), Min=min(x), Mean=mean(x), N=length(x))
d[, g(bhr), by=.(gender, newMI)]  # if g returned a vector instead, use as.list(g(bhr))

d[, as.list(quantile(bhr)), by=gender]

# Compute mean bhr by quintiles of age using Hmisc::cut2
# Bad statistical practice; use scatterplot + smoother instead
d[, .(Mean=mean(bhr)), keyby=.(fifth=cut2(age, g=5))]

# Compute multiple statistics on multiple variables
d[, lapply(.SD, quantile), by=gender, .SDcols=.q(bhr, pkhr, sbp)]
# Similar but put percentile number in front of statistic value
# Do only quartiles
g <- function(x) {
  z <- quantile(x, (1:3)/4, na.rm=TRUE)
  paste(format(names(z)), format(round(z)))
}
d[, lapply(.SD, g), by=gender, .SDcols=.q(bhr, pkhr, sbp)]
# To have more control over labeling and to have one row per sex:
g <- function(x) {
  s <- sapply(x, quantile, na.rm=TRUE)  # compute quantiles for all variables -> matrix
  h <- as.list(s)  # vectorizes first
  # Cross row names (percentile names) with column (variable) names
  # paste(b, a) puts variable name in front of percentile
  names(h) <- outer(rownames(s), colnames(s), function(a, b) paste(b, a))
  h
}
d[, g(.SD), by=gender, .SDcols=.q(bhr, pkhr, sbp)]
# Restrict to variables bhr - basedp in order columns created in data table
d[, g(.SD), by=gender, .SDcols=bhr : basedp]
# Can put ! in front of a sequence of variables to do the opposite

# To add duplicated means to raw data use e.g.
# d[, Mean := mean(x), by=sex]
```

## Summary Statistics Using Functions Returning Two-Dimensional Results {#sec-sstats-mult}

`data.table` can create objects where results are matrices instead of data tables, or where table elements are not just scalar quantities but are `list`s.  We previously saw examples where the summarization function produced a vector or where complex side effects happened without storing the results in a data table.  Now consider examples where the summarization function produces a matrix or a list of lists and the final result is a matrix or a more complex data table.  Start with calculating three quartiles on one variable two different ways.  The examples that follow do not give satisfactory labeling when a `by` variable is added.

```{r}
qu <- function(x, type=7) {
  y <- quantile(x, (1:3)/4, na.rm=TRUE, type=type)
  names(y) <- .q(Q1, Q2, Q3)
  y
}
g  <- function(x) rbind(type7=qu(x), type8=qu(x, type=8))
d[, g(pkhr)]   # 2 x 3 matrix
# d[, g(pkhr), by=gender] does not retain names
```

Compute 3 quartiles for each of 3 variables.

```{r}
g <- function(x) lapply(x, qu)
d[, g(.SD), .SDcols=.q(bhr, pkhr, sbp)]  # a data.table w/out Q1 Q2 Q3 labels
g <- function(x) apply(x, 2, qu)
d[, g(.SD), .SDcols=.q(bhr, pkhr, sbp)]  # a matrix with Q1 Q2 Q3 labels
```

Now consider a summarization function computing a 2 $\times$ 2 result where some values are numeric and some are character strings.  Separately for each value of a `by` variable we desire the name of the variable having the lowest number of missing values, and the frequency of `NA`s, and the name of the variable having the highest number of missing values, and that frequency.  We use the `pbc` dataset that comes with the `survival` package, and do the variable search separately by outcome status.  The summarization function `g` operates on a data table for the current outcome status.  Consider only the variables in `pbc` that are ever `NA`.

```{r}
p <- copy(pbc)
setDT(p, key='status')  # sort by status
numna <- sapply(p, function(x) sum(is.na(x)))  # loops over vars and counts NAs -> vector
v     <- names(numna[numna > 0])   # names of variables with any NA 

g <- function(x) {
  m <- sapply(x, function(y) sum(is.na(y)))
  mn <- m[which.min(m)]   # keeps variable name in names attribute
  mx <- m[which.max(m)]
  list(Lowest  = list(VariableLow  = names(mn), Nlow  = mn),
       Highest = list(VariableHigh = names(mx), Nhigh = mx))
}
w=p[, g(.SD), by=status, .SDcols = v]   # data.table
w
sapply(w, class)
w[1]
w[1, Lowest]
w[2, Lowest]
```

The `Lowest` and `Highest` elements of a given row of `w` are R `list`s, allowing for mixed object types (character/integer).  This doesn't make for the most concise way to retrieve individual elements but `w` prints well.  We can make a more conventional data table with the following code.

```{r}
g <- function(x) {
  m <- sapply(x, function(y) sum(is.na(y)))
  mn <- m[which.min(m)]
  mx <- m[which.max(m)]
  list(VarLow  = names(mn), Nlow  = mn,
       VarHigh = names(mx), Nhigh = mx)
}
w=p[, g(.SD), by=status, .SDcols = v]   # data.table
sapply(w, class)
w
```

## Summary Statistics With Marginal Summaries {#sec-sstats-marg}

The `cube` function in the `data.table` package will compute all possible marginal statistics.  When there is only one `by` variable, the overall statistic is computed in addition to compute stratified values.  When a dimension is being marginalized over, the value of the `by` variable for that dimension will be `NA`.

```{r cube}
mn  <- function(x) as.double(mean(x, na.rm=TRUE))
# as.double ensures consistency of numeric type across groups
Nna <- function(x) sum(! is.na(x))
cube(d, .(Meanbhr = mn(bhr), N = Nna(bhr)), by='gender', id=TRUE)
cube(d, .(Meanbhr = mn(bhr), N = Nna(bhr)), by=.q(gender, hxofMI), id=TRUE)
# id=TRUE creates output variable grouping to detail level of marginalization
# It is a binary representation, e.g. if by has 3 variables and a row
# is marginalizing over the first and third variables,
# grouping=binary 101 = 5
# Use groupingsets() to control the marginalizations
# Example: marginalize only one variable at a time
groupingsets(d, .(Meanbhr = mn(bhr), N=Nna(bhr)), by=.q(gender, hxofMI),
             sets=list('gender', 'hxofMI'), id=TRUE)
```